---
title: "Custom Families"
author: "Robin Evans"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
options(digits=3)
```

The `causl` package now allows users to specify custom families with which to
perform their simulations.  To do this, one must create a function that takes as
an argument the link function, and returns a list containing:

 * `name`: the name of the relevant family;
 * `ddist`: a function returning the density of the distributions;
 * `qdist`: a function returning the quantiles from probabilities;
 * `rdist`: a function to sample values from the distribution;
 * `pdist`: a cumulative distribution function;
 * `pars`: a list of the names of the parameters used;
 * `default`: a function that returns a list of the default values for an 
 observation and each of the parameters;
 * `link`: the specified link function.
 
The function should also give the output the class `"causl_family"`, so that
it is interpreted appropriately.
 
 
## Example (Poisson distribution)

Suppose we wish to simulate from a Poisson distribution as part of a frugal 
model.  Then we can define a function such as the below:
 
```{r pois, message=FALSE}
library(causl)

poisson_causl_fam <- function (link) {
  if (missing(link)) link = "log"

  ## write functions
  dens <- function (x, mu, log=FALSE) dpois(x, lambda=mu, log=log)
  quan <- function (p, mu) qpois(p, lambda=mu)
  sim <- function (n, mu) {
    qx <- runif(n)
    x <- qpois(qx, lambda=mu)
    attr(x, "quantile") <- qx
    return(x)
  }
  probs <- function (x, mu) ppois(x, lambda=mu)

  default <- function(theta) list(x=1, mu=1)

  ## define family
  out <- list(name="poisson", ddist=dens, qdist=quan, rdist=sim, pdist=probs,
              pars=c("mu"), default=default, link=link)
  class(out) <- "causl_family"

  return(out)
}
```
Some remarks are in order.  Note that the `ddist` function must have a `log` 
argument to allow the log-density to be evaluated.  In addition, since this
is a discrete distribution, if we use it in a copula we must know the exact 
quantile used in the simulation (not just the one returned by applying 
`ppois`).  Hence we simulate a uniform and then obtain the Poisson random
variables by inversion from `qpois`.  For continuous distributions this is 
unnecessary, and `glm_sim` will just apply the `pdist` function. 

Now, let us try to simulate some data using this distribution.  We will need
to use `poisson_causl_fam` in our `family` argument:
```{r params}
forms <- list(Z ~ 1, X ~ Z, Y ~ X, ~ 1)
fams <- list(1, poisson_causl_fam(), 1, 1)
pars <- list(Z = list(beta = 0, phi = 1),
             X = list(beta = c(0.3,1)),
             Y = list(beta = c(-1,0.5), phi = 1),
             cop = list(beta = 1))
cm <- causl_model(formulas = forms, family = fams, pars = pars)

set.seed(123)
dat <- rfrugal(n=1e3, causl_model=cm)
```

### Treatment model

Then we can check that the new variables have been simulated correctly.  We 
ought to have:
\begin{align*}
P(X = x \mid z) &= \frac{1}{x!} \exp\{\alpha_0 + \alpha_1 z x - e^{\alpha_0 + \alpha_1 z}\}, 
\end{align*}
with $\alpha = `r pars$X$beta[1]`$ and $\alpha_1 = `r pars$X$beta[2]`$.
Let's perform a regression to check the parameters:
```{r chk_pois}
modX <- glm(X ~ Z, family=quasipoisson, data=dat)
summary(modX)$coefficients
```
So we can see that the estimates are well within two standard errors of the 
true value.  In addition, 
```{r chk_pois_disp}
summary(modX)$dispersion
```
the dispersion parameter is close to its nominal value of 1.

### Outcome model

We can also fit the outcome model using maximum likelihood estimation with 
sandwich errors.
```{r fit_causl}
out <- fit_causl(dat = dat, formulas = list(Y ~ X, Z ~ 1, cop ~ 1),
                 family = c(1, 1, 1))
```
This gives us estimates of $\beta_0 = `r out$par[1]`$ (`r out$sandwich_se[1]`) and
$\beta_1 = `r out$par[2]`$ (`r out$sandwich_se[2]`), which is consistent with 
the values $-1$ and $0.5$ that we used to simulate.

<!-- We can also attempt to infer the causal effect of $X$ on $Y$: -->
<!-- ```{r fitX} -->
<!-- eta <- c(cbind(1,dat$Z) %*% modX$coefficients) -->
<!-- gps <- dpois(dat$X, lambda=exp(eta)) -->
<!-- mean((dat$Y/gps)*(dat$X==1)) - mean((dat$Y/gps)*(dat$X==0)) -->
<!-- mean((dat$Y/gps)*(dat$X==2)) - mean((dat$Y/gps)*(dat$X==1)) -->
<!-- mean((dat$Y/gps)*(dat$X==3)) - mean((dat$Y/gps)*(dat$X==2)) -->
<!-- mean((dat$Y/gps)*(dat$X==4)) - mean((dat$Y/gps)*(dat$X==3)) -->
<!-- ``` -->

## Custom Link Functions

The example above works because `log` is an existing link function specified
in the package.  Suppose, however, that we want to use the `sqrt` function as
the link; this is not used with any built-in families. 

For this we need to specify a `custom_links` argument to `poisson_causl_fam`; 
each entry in this named list should itself be a list, containing:

 * `linkfun`: the link function $g$;
 * `linkinv`: the inverse of $g$.
 
The names of the entries should correspond to the string used as the name of 
the link function.
```{r pois2}
 poisson_causl_fam <- function (link) {
  if (missing(link)) link = "log"

  ## write functions
  dens <- function (x, mu, log=FALSE) dpois(x, lambda=mu, log=log)
  quan <- function (p, mu) qpois(p, lambda=mu)
  sim <- function (n, mu) {
    qx <- runif(n)
    x <- qpois(qx, lambda=mu)
    attr(x, "quantile") <- qx
    return(x)
  }
  probs <- function (x, mu) ppois(x, lambda=mu)

  default <- function(theta) list(x=1, mu=1)
  
  custom_links <- list(sqrt = list(linkfun = function (x) sqrt(x),
                                   linkinv = function (x) x^2))

  ## define family
  out <- list(name="poisson", ddist=dens, qdist=quan, rdist=sim, pdist=probs,
              pars=c("mu"), default=default, link=link, custom_links=custom_links)
  class(out) <- "causl_family"

  return(out)
}
```
Now we can suitably modify our earlier example:
```{r params_sqrt}
fams <- list(1, poisson_causl_fam(link="sqrt"), 1, 1)
pars$X$beta <- c(0.9, 0.05)
cm_sq <- modify.causl_model(cm, family=fams, pars=pars)

set.seed(124)
dat <- rfrugal(1e3, causl_model = cm_sq, control=list(quiet=TRUE))
```
We can again check that the distribution was correctly sampled.
```{r chk_pois_sqrt}
modX <- glm(X ~ Z, family=quasipoisson(sqrt), data=dat, start=c(1,0))
summary(modX)$coefficients
```
So we again see that the estimates are (just about) within two standard errors 
of the true values (0.9 and 0.05), and the dispersion parameter is estimated to 
be `r summary(modX)$dispersion`, again close to 1.

