---
title: "Discrete Variables Copula"
author: "Robin J. Evans"
date: "`r Sys.Date()`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Discrete Variables Copula}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=3)
library(causl)
library(survey)
```

## Incorporating discrete variables

We can also fit models in which some of the covariates and/or outcomes are 
binary.  For example, suppose we assume that 
\begin{align*}
Z_1 &\sim N(0, \sigma_1^2) & Z_2 &\sim \text{Bernoulli}(p_2) & Y \mid do(X=x) \sim N(0, \sigma_y^2)
\end{align*}
and $X \mid Z_1=z_1,Z_2=z_2 \sim \text{Bernoulli}(q)$, and we assume a Gaussian
copula over $Z_1,Z_2,Y$ with correlation $R$, where $\rho_{Z_1Z_2} = 0.5$, 
$\rho_{Z_1Y} = 0.3$ and $\rho_{Z_2Y} = 0.4$.

Then we can set up this model in the usual way:
```{r settings}
forms <- list(list(Z1 ~ 1, Z2 ~ 1), X ~ Z1*Z2, Y ~ X, ~ 1)
fams <- list(c(1,5), 5, 1, 1)
pars <- list(Z1 = list(beta=0, phi=1),
             Z2 = list(beta=0),
             X = list(beta=c(-0.3,0.1,0.2,0)),
             Y = list(beta=c(-0.25, 0.5), phi=0.5),
             cop = list(beta=matrix(c(0.5,0.3,0.4), nrow=1)))
```
Now call `rfrugalParam` as usual.  
```{r simulate}
set.seed(1234)
n <- 1e4
dat <- rfrugalParam(n, formulas = forms, family = fams, pars = pars)
```
Then we can check that the distributions follow the specification that we 
asked for.
```{r check_disc}
ks.test(dat$Z1, pnorm)
binom.test(sum(dat$Z2), n=n, p = 0.5)
glmX <- glm(X ~ Z1*Z2, family=binomial, data=dat)
summary(glmX)$coefficients
```
These are all consistent with the values we provided.  We can finally check the
outcome model.
```{r check_Y}
ps <- predict(glmX, type="response")
wt <- dat$X/ps + (1-dat$X)/(1-ps)
glmY <- svyglm(Y ~ X, design = svydesign(~1, weights=wt, data=dat))
summary(glmY)$coefficients
```

## Categorical and Ordinal Variables

We can also extend this to variables that are categorical or ordered 
categorical.  These options correspond to the family indicators 11 and 10 
respectively, or they can be accessed by passing the strings `"categorical"` 
and `"ordinal"` to the `family` argument.  Both types of variable are stored as
a factor, with labels for the objects `1`, `2`, up to $\ell$, where $\ell$ is
the number of levels.  

These methods respectively use a 
contrast with the baseline level, or a contrast with the category immediately 
below the current one in the ordering.  For example, if we have a three variable
categorical variable using the formula `Z ~ A` then the parameters are of the 
form:
\begin{align*}
\log \frac{P(Z=i)}{P(Z=1)} &= \beta_{0i} + \beta_{ai} a, & & i = 2,\ldots,\ell.
\end{align*}
For a categorical variable the contrasts become $\log \{P(Z=i)/P(Z=i-1)\}$ 
for the same range of $i$. 
 
The regression parameters can be passed either as a vector or a matrix, but in
either case they must be ordered so that all the coefficients for one contrast
precede all those for another.  For example, if we want we could put
`Z = list(beta = c(0.5,0.1,-0.5,0.4), nlevel = 3)` in the `pars` argument to 
require that 
\begin{align*}
\log \frac{P(Z=2)}{P(Z=1)} &= 0.5 + 0.1 a & \log \frac{P(Z=3)}{P(Z=1)} &= -0.5 + 0.4 a
\end{align*}
for an unordered categorical variable, and the same for an ordinal variable but 
replacing the second quantity with $\log \{P(Z=3)/P(Z=2)\}$.  Equivalently, we 
could put `Z = list(beta = matrix(c(0.5,0.1,-0.5,0.4), nrow=2), nlevel = 3)`, 
and the same model would be fitted.

```{r}
forms <- list(list(Z0 ~ A0, Z1 ~ A0),
              list(A0 ~ 1, A1 ~ A0*Z0),
              Y ~ A0*A1,
              ~ A0)
fams <- list(c("categorical","categorical"),c(5,5),1,1)
pars <- list(A0 = list(beta = 0),
             Z0 = list(beta = c(0.3,-0.2,0.4,0.1), nlevel=3),
             Z1 = list(beta = c(0.3,-0.2,0.4,0.1), nlevel=3),
             A1 = list(beta = 2*c(-0.3,0.4,0.3,0.5,0.3,0.5)),
             Y = list(beta = c(-0.5,0.2,0.3,0), phi=1),
             cop = list(beta=c(2,0.5)))
```
In this case we set both covariate variables to be categorical. 
```{r}
set.seed(123)
n <- 5e4
dat <- rfrugalParam(n, formulas = forms, pars=pars, family = fams)
```
Now we can check that the parameters being simulated match those input.
```{r check_cat}
glm(I(Z0==2) ~ A0, family=binomial, data=dat[dat$Z0 != 3,])$coefficients
glm(I(Z0==3) ~ A0, family=binomial, data=dat[dat$Z0 != 2,])$coefficients
glm(I(Z1==2) ~ A0, family=binomial, data=dat[dat$Z1 != 3,])$coefficients
glm(I(Z1==3) ~ A0, family=binomial, data=dat[dat$Z1 != 2,])$coefficients 
```
Finally, we can check that the outcome model works as it should.
```{r check_Yc}
ps <- predict(glm(A1 ~ A0*Z0, family=binomial, data=dat), type="response")
wt <- dat$A1/ps + (1-dat$A1)/(1-ps)
summary(svyglm(Y ~ A0*A1, design = svydesign(~1,data=dat,weights=wt)))$coef
```
Comparing with a naïve (unweighted) estimate.
```{r naive_c}
summary(svyglm(Y ~ A0*A1, design = svydesign(~1,data=dat,weights=~1)))$coef
```
