---
title: "Discrete Variables Copula"
author: "Robin J. Evans"
date: "`r Sys.Date()`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Discrete Variables Copula}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Incorporating discrete variables

We can also fit models in which some of the covariates and/or outcomes are 
binary.  For example, suppose we assume that 
\begin{align*}
Z_1 &\sim N(0, \sigma_1^2) & Z_2 &\sim \text{Bernoulli}(p_2) & Y \mid do(X=x) \sim N(0, \sigma_y^2)
\end{align*}
and $X \mid Z_1=z_1,Z_2=z_2 \sim \text{Bernoulli}(q)$, and we assume a Gaussian
copula over $Z_1,Z_2,Y$ with correlation $R$, where $\rho_{Z_1Z_2} = 0.5$, 
$\rho_{Z_1Y} = 0.3$ and $\rho_{Z_2Y} = 0.4$.

Then we can set up this model in the usual way:
```{r settings}
forms <- list(list(Z1 ~ 1, Z2 ~ 1), X ~ Z1*Z2, Y ~ X, ~ 1)
fams <- list(c(1,5), 5, 1, 1)
pars <- list(Z1 = list(beta=0, phi=1),
             Z2 = list(beta=0),
             X = list(beta=c(-0.3,0.1,0.2,0)),
             Y = list(beta=c(-0.25, 0.5), phi=0.5),
             cop = list(beta=matrix(c(0.5,0.3,0.4), nrow=1)))
```

```{r simulate}
set.seed(123)
n <- 1e4
dat <- rfrugalParam(n, formulas = forms, family = fams, pars = pars)
```
