---
title: "Hidden Variables"
author: "Robin J. Evans"
date: "27/05/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hidden Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this example we use the `survey` package to get robust standard errors when 
reweighting.  We start by loading the libraries.
```{r setup, echo=2:3, message=FALSE}
knitr::opts_chunk$set(echo = TRUE); options(digits=3)
library(causl)
library(survey)
```

## Set Up the Model

We begin by setting up the formulas, families and parameter values:
```{r setup_model}
formulas = list(list(U ~ 1, L ~ A0), list(A0 ~ 1, A1 ~ A0*L), Y ~ A0*A1, ~A0*A1)
fam = list(c(4,3), c(5,5), c(3), c(1,1,1))

pars <- list(A0 = list(beta = 0),
             U = list(beta = 0, phi=1),
             L = list(beta = c(0.3,-0.2), phi=1),
             A1 = list(beta = c(-0.3,0.4,0.3,0)),
             Y = list(beta = c(-0.5,0.2,0.3,0), phi=1),
             cop = list(beta = matrix(c(1,0,0,0,
                                 1,0,0,0,
                                 0.5,0,0,0), nrow=4)))
```

```{r simulate}
set.seed(125)
n <- 1e4
dat <- causalSamp(n, formulas, family=fam, pars=pars)
```
We can then check that the parameter values match their intended values:
```{r check_data}
summary(svyglm(L ~ A0, family=Gamma(link="log"), 
               design=svydesign(id=~1, weights=~1, data=dat)))$coef
glmA1 <- glm(A1 ~ A0*L, family=binomial, data=dat)
summary(glmA1)$coef
w <- predict(glmA1, type="response")
dat$wt <- dat$A1/w + (1-dat$A1)/(1-w)

summary(svyglm(Y ~ A0*A1, family=Gamma(link="log"), 
               design=svydesign(id=~1, weights=~wt, data=dat)))$coef
```
Indeed, they are all within two standard errors of their nominal values.
```{r, echo=FALSE, eval=FALSE}
# summary(glm(Y ~ A0*A1, family=Gamma(link="log"), weights = wt, data=dat))$coef
```

We can also fit the data using maximum likelihood directly.
```{r fit_data, eval=FALSE}
out <- fitCausal(dat, formulas = list(L ~ A0, Y ~ A0*A1, ~1), family = c(3,3,1))
out
```
Again, all estimates are within two standard errors of the true values.

