---
title: "Inversion Tutorial"
author: "Robin J. Evans"
date: "10/06/2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hidden Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this tutorial, we show how to simulate using the inversion method, which is
significantly faster than rejection sampling.  As always, start by loading the 
package as well as the `survey` package.
```{r setup, echo=2:3, message=FALSE}
knitr::opts_chunk$set(echo = TRUE); options(digits=3)
library(causl)
library(survey)
```


## Set Up the Model

We begin by setting up the formulas, families and parameter values:
```{r setup_model}
formulas = list(list(U ~ 1, L ~ U*A0),   # covariates
                list(A0 ~ 1, A1 ~ A0*L), # treatments
                Y ~ A0*A1,             # outcome
                ~ A0*A1)               # copula
fam = list(c(4,3), c(5,5), c(3), c(1,1,1))

pars <- list(A0 = list(beta = 0),
             U = list(beta = 0, phi=1),
             L = list(beta = c(0.3,0.5,-0.2,-0.1), phi=1),
             A1 = list(beta = c(-0.3,0.4,0.3,0)),
             Y = list(beta = c(-0.5,0.2,0.3,0), phi=1),
             cop = list(beta = list(list(c(1,0,0,0), c(0.5,0,0,0)))) )
```

```{r simulate}
set.seed(123)
n <- 1e4
dat <- rfrugalParam(n, formulas, family=fam, pars=pars, method="inversion")
# dat <- causalSamp(n, formulas, family=fam, pars=pars)
```
We can then check that the parameter values match their intended values:
```{r check_data}
summary(glm(L ~ U*A0, family=Gamma(link="log"), data=dat))$coef
glmA1 <- glm(A1 ~ A0*L, family=binomial, data=dat)
summary(glmA1)$coef
w <- predict(glmA1, type="response")
dat$wt <- dat$A1/w + (1-dat$A1)/(1-w)

## wrong model
mod_w <- svyglm(Y ~ A0*A1, family=Gamma(link="log"), 
                design = svydesign(~1, weights=rep(1,nrow(dat)), data=dat))
mod_w2 <- glm(Y ~ A0*A1, family=Gamma(link="log"), data=dat)
summary(mod_w)$coef
## correct model
mod_c <- svyglm(Y ~ A0*A1, family=Gamma(link="log"), 
                design = svydesign(~1, weights=~wt, data=dat))
summary(mod_c)$coef
```
The `A0` and `A1` coefficients of the naive model and the correct
model show IPW works very well.

