---
title: "Plasmode Simulation"
author: "Robin Evans"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=3)
```

## Introduction

This document describes how to use the `causl` package to perform plasmode 
simulation; that is, producing datasets that combining real covariates (and 
possibly treatments) with synthetic outcomes.  This enables us to have realistic
datasets, but also to have knowledge of the underlying causal effect, making 
them very useful for testing the relative effectiveness of different inference
methods.

Start by loading the `causl` and `dplyr` packages.

```{r cars, message=FALSE}
library(dplyr)
library(causl)
```

## Dataset

We will use the dataset from a competition held at the Atlantic Causal Inference 
Conference in 2016.  To obtain this, first run the following commands:
```{r install_dorie, eval=FALSE}
install.packages("remotes")
remotes::install_github("vdorie/aciccomp/2016")
```
Then load the `aciccomp2016` package.
```{r load_dorie}
library(aciccomp2016)
(dat <- as_tibble(input_2016)) # show 10 rows of first few variables
```

## Model

Let us consider a model for the causal effect of smoking on birthweight.  We will
start with a binary variable $A$ to indicate whether the mother smoked during the
pregnancy, and later extend this to a zero-inflated continuous variable.  We
start by setting up the model inputs:

```{r setup1}
forms <- list(list(),
              list(A ~ x_1 + x_3 + x_4),
              list(Y ~ A),
              list(~ 1))
fams <- list(integer(0), 5, 1, 1)
pars <- list(A = list(beta=c(-1.5,0.03,0.02,0.05)),
             Y = list(beta=c(3200, -500), phi=400^2),
             cop = list(beta=-1))
```
Then call `rfrugalParam` to simulate $A$ and $Y$.
```{r simdat1}
datAY <- rfrugalParam(formulas=forms, family=fams, pars=pars, dat=dat)
```
We can now check that this basic simulation was performed correctly.  First we
fit the treatment variable using the correct model.
```{r chkA1}
glmA <- glm(A ~ x_1 + x_3 + x_4, family=binomial, data=datAY)
summary(glmA)$coefficients
```
Indeed, the parameters appear correct.  Then we can use inverse weighting to
estimate the parameters for the outcome model.  We will need to load the 
`survey` package to get robust standard errors after the weighting.
```{r chkY1, message=FALSE, warning=FALSE}
library(survey)
ps <- predict(glmA, type="response")
wt <- datAY$A/ps + (1-datAY$A)/(1-ps)
glmY <- svyglm(Y ~ A, design = svydesign(~1, weights = wt, data=datAY))
summary(glmY)$coef
```

We can also compare with a naïve method that ignores reweighting.

```{r chkY1n, message=FALSE, warning=FALSE}
glmYn <- glm(Y ~ A, data=datAY)
summary(glmYn)$coef
```
